---
title: "Gym Attendance RCT Data Cleaning and Regression Analysis"
author: "Georgianna James"
date: "October 20th, 2021"
output: github_document
---


# Set up 

## Required Packages 



```{r packages, message=FALSE}
library(tidyverse)
library(ggplot2)
library(stringr)
library(plotrix)
library(here)
library(stargazer)
```

## Import Data



```{r message=FALSE}
raw_data <- read_csv(here("data", "hoursatgymRCT.csv"))

gym_attendance <- read_csv(here("data", "hoursatgymRCT.csv"))

```

## Calculating Average Treatment Effect before cleaning, with no controls 

```{r echo=FALSE}

regression <- lm(hours~treatment, data = gym_attendance)



stargazer(regression,
           type = "text",
  title="Effects of Treatment (Compensation for attending gym)", 
  dep.var.labels=c("Hours"),
  covariate.labels=c("Treatment","Control Mean"),
  digits=3, out = "table1.txt")
```




After running a regression, I found an average treatment effect and standard error as follows:

Average Treatment Effect = 6.594
Standard Error = 5.992

This regression is not statistically significant, with a p value of 0.2714. However, I must clean the data and add controls to determine whether or not this relationship is truly statistically significant. 


# Data Cleaning 

## Hours Cleaning

After 30 hours, the hour distribution drops off until around 60 hours and then forms a distribution of a similar shape. I assume that this is because these seemingly unreasonable hour reportings were reported in minutes, so I am going to convert them back to hours by dividing by 60. 

```{r }
gym_attendance <- gym_attendance %>% 
  mutate(hours = if_else(hours > 30, hours / 60, hours))
```


## Community Center Cleaning

The community center column is not organized. There are a variety of inputs for both hyde park and woodlawn, when we only have 2 variables. I am going to change all hyde park variables to "H" and all Woodlawn variables to "W" so that there are only two community center groups. 

```{r }
## clean community center
gym_attendance <- 
  gym_attendance %>%
  mutate(`community_center` = substr(`community_center`,1,1),
         `community_center` = toupper(`community_center`))
      
```

## Gender Cleaning

I am going to change all "female" to 1 and all "male" to 0, creating a female dummy variable.

```{r }
gym_attendance <-
  gym_attendance %>%
  mutate(`female` = str_replace(`female`, c("female"),"1"),
         `female` = str_replace(`female`, "male", "0"),
         `female` = as.double(`female`))
```

## Age Cleaning


I am going change all of the participants whose age is reported as -99 with NA. 




```{r }
gym_attendance <- gym_attendance %>% 
  mutate(age = na_if(age, -99))

```


## BMI Cleaning




```{r }
gym_attendance <- gym_attendance %>% 
  mutate(bmi = if_else(bmi < 1, bmi * 100, bmi))
```


`

Because these outliers form a similar shaped distribution to the rest of the bmi data and BMI typically falls between 15% and 30%, I assume all the recorded BMI values that are less than 1 were actually bmi's reported as percentages, so I changed these from decimal percentages to numbers by multiplying by 100. For example 0.22 becomes 22. 



## Race_ethnicity cleaning

There are two different black variables. Assuming that they mean the same thing. I am going to change all black variables to "B." For continuity I will change hispanic to "H", and white to "W". 

```{r }
gym_attendance <- 
  gym_attendance %>%
  mutate(`race_ethnicity` = substr(`race_ethnicity`,1,1),
         `race_ethnicity` = toupper(`race_ethnicity`))

```




## Regressing with clean data

```{r echo=TRUE}
regression <- lm(hours~treatment, data = gym_attendance)




stargazer(regression,
           type = "text",
  title="Effects of Treatment (Compensation for attending gym)", 
  dep.var.labels=c("Hours"),
  covariate.labels=c("Treatment","Control Mean"),
  digits=3, out = "table2.txt")

```



After running a regression with the cleaned data, I found a treatment effect and standard deviation of:

average treatment effect: 1.8740
standard deviation: 0.2233

While these results are significant, I must to ensure there was true randomization before I can interpret these results. 


## Evaluating Randomization

In order to have true randomization, the treatment and control groups must be otherwise identical to each other. Additionally, there must not be any selection bias in your population that may come from self selection into the group, non-compliance, attrition, and many other confounding variables. In order to analyze what went wrong within this randomization, I must evaluate how the covariates influence the output of each group. 




```{r echo=FALSE}
gym_attendance %>% 
  group_by(treatment) %>% 
  summarise(average_age = mean(age, na.rm = TRUE), average_bmi = mean(bmi), average_hours = mean(hours))
```


Looking at race distribution across treatment and control:

```{r echo=FALSE, message=FALSE, warning=FALSE}
gym_attendance %>% 
  group_by(treatment, race_ethnicity) %>% 
  summarize(count = n())
```


The control group has 194 black participants, 206 white participants, and 86 hispanic participants, and the treated group has 315 black participants, 120 white participants, and 51 hispanic participants, which are similar distributions. 

Now, let's look at community center distributions:



```{r echo=FALSE, message=FALSE, warning=FALSE}
gym_attendance %>% 
  group_by(treatment, community_center) %>% 
  summarize(count = n(), average_hours = mean(hours), average_bmi  = mean(bmi), average_age = mean(age, na.rm = TRUE))
```


The treated group had 350 at Woodlawn and 150 at Hyde Park. The distribution between community center is very imbalanced between control and treatment. Additionally, the type of patrons who attend the two different commmunity centers differ by average bmi and age. 
 

```{r include=FALSE}
regression <- lm(hours~community_center, data = gym_attendance)

summary(regression)
```


```{r echo=FALSE}
stargazer(regression,
           type = "text",
  title="Effects of Community Center (Compensation for attending gym)", 
  dep.var.labels=c("Hours"),
  covariate.labels=c("Woodlawn","Intercept"),
  digits=3, out = "table1.txt")
```

After running a regression I found that simply participating in this experiment at Woodlawn had a large, positive, statistically significant impact on hours reported. The effect of being at Woodlawn on hours was 2.674. The treatment group had 350/500 participants at Woodlawn while the control group had 150/500. This is a very significant confounding variale. While mediators such as convenience, accessibility, or proximity to facility may be influencing this relationship, it was captured by the community center variable.

```{r echo=FALSE}
gym_attendance %>% 
  group_by(treatment) %>% 
  summarise(count = n(), female = sum(female), percentfemale = sum(female)/count)
```
There are different distributions of gender in treatment and control. 




## Controling for Community Center and Gender

In order to know the true treatment effect, you must control for community center and gender. Running a regression, controlling for community center and gender, you find: 


```{r include=FALSE}
regression <- lm(hours~treatment + community_center + female, data = gym_attendance)


summary(regression)
```

```{r echo=FALSE}
stargazer(regression,
                     type = "text",
  title="Effects of Treatment (Controlling for community center and gender)", 
  dep.var.labels=c("Hours"),
  covariate.labels=c("Treatment","Community Center", "Female","Control Mean"),
  digits=3, out = "table3.txt"
  )

```











